### 2025-06-02 manuscript/main.tex
dsl, paradigms

The terms \textit{abstract} and \textit{concrete} syntax are primarily encountered in the context of programming language design. While the two concepts follow the common relationship of
abstract and concrete instances — where one embodies a generalized, meta-level blueprint, while the other encapsulates the concrete, instance-level details — , and the term "syntax" broadly applies to all languages,
the idea of pinning meta-information onto language constructs, somewhat implies further calculations based on the language itself.

@@

Languages used for further computation can generally be slotted into two overarching paradigms, \textbf{imperative} and \textbf{declarative} languages, ...
in the way of reaching the desired outcome.
Declarative languages describe \textit{what} the computation should accomplish by adhering to sets of expressions, constraints or logical statements. \cite{9,10}
These formulas set the rules and goals characterizing the desired result or relationship of input and output values. The actual control-flow and low-level decisions of
the execution is left to the specific language implementations and runtime engines.
The paradigm terms themselves a rather vague adn leave room for interpretation?. It is also the case that many languages do not purely fit into one or the other and lend
part of their ideas to each other. Eg. haskell, .NET, ...

Imperative more verbose but low level access...

\begin{itemize}
  \item **Imperative**: describe *how* to do things (sequential steps, mutable state, control flow).
  \item **Declarative**: describe *what* you want (constraints, relationships, goals), leaving execution order to the runtime or solver.
  \item Key trade-offs: control vs.\ abstraction, predictability vs.\ conciseness, ease of optimization, typical use-cases.
\end{itemize}

\section{Domain-Specific Languages (DSLs)}
\begin{itemize}
  \item Definition and motivation: tailor syntax/semantics to a narrow problem domain.
  \item Classification by paradigm:
    \begin{itemize}
      \item **Imperative DSLs** (e.g.\ MiniProb, CUDA, many build systems): embed domain logic in a procedural style.
      \item **Declarative DSLs** (e.g.\ SQL, HTML, many configuration languages): express constraints or data relationships.
    \end{itemize}
  \item Criteria for choosing one over the other: domain control, ease of reasoning, tool support, typical analysis patterns.
\end{itemize}

\subsection{Imperative DSLs (Your Thesis Focus)}
\begin{itemize}
  \item Overview of their structure: explicit state, control flow, side-effects.
  \item Example: MiniProb (POPACheck)  
    \begin{itemize}
      \item Recursive calls, `observe`, mutable grid updates.  
      \item How it compiles down to a probabilistic push-down automaton.  
    \end{itemize}
  \item Pros \& Cons in your domain:  
    \begin{itemize}
      \item Pro: fine-grained control, easy to encode sequential reasoning.  
      \item Con: harder to optimize globally, side-effects can obscure analysis.  
    \end{itemize}
  \item Comparison with Declarative DSLs for the same problem (e.g.\ PCTL, PRISM models).  
\end{itemize}

\subsection{Declarative DSLs}
\begin{itemize}
  \item What makes a DSL declarative: absence of explicit control flow, emphasis on logical relations.  
  \item Typical examples in probabilistic modeling: PCTL, PRISM, probabilistic logic programs.  
  \item Why—despite their advantages—MiniProb remained imperative in POPACheck.  
\end{itemize}

DSL or General (do api prog lang have to be haskell complete? C3, java haskell ->>)

>>>>

The terms \textit{abstract} and \textit{concrete} syntax are primarily encountered in the context of programming language design. While the two concepts follow the common relationship of
abstract and concrete instances — where one embodies a generalized, meta-level blueprint, while the other encapsulates the concrete, instance-level details — and the term "syntax" broadly applies to all languages,
the idea of pinning meta-information onto language constructs, somewhat implies further calculations based on the language itself.

@@

Languages used for further computation can be analyzed from multiple perspectives and classified with different qualities in mind. \cite{progParadigmsForDummiess}
For this thesis we focus on slotting languages into two overarching paradigms, \textbf{imperative} and \textbf{declarative} languages, 
which differentiate the  by inspecting way of reaching the desired outcome.?NOLIKETHIS?
Declarative languages describe \textit{what} the computation should accomplish by adhering to sets of expressions, constraints or logical statements. \cite{grammarDrivenDSLDebug}
These formulas set the rules and goals characterizing the desired result or relationship of input and output values. The actual control flow and low-level decisions of
the execution is left to the specific language implementations and runtime engines.

Generally, because of the vague wording used, the paradigms themselves are not always completely and accurately describing the behaviour of a programming language,
possibly leading to different concepts overlapping. This difference is especially clear in modern languages, which often implement functionality by borrowing 
constructs from the opposite paradigm — for example, imperative C\# incorporates declarative features through its LINQ-Library, while Haskell, functional by 
design, supports imperative-style programming. \cite{netHaskellClaims}\\

Imperative languages, in contrast to declarative ones, describe \textit{how} a computation will accomplish the desired end-state by following explicit sequences
of commands. These commands consist of statements that actively manipulate the computations state, which is typically managed through assignments to variables 
and taking charge of the control flow with provided language constructs (loops, conditionals, calls, procedures, etc.).
Consequently, correct execution is no longer responsibility of the underlying system, but the implementation and therefore the programmer itself. The imperative paradigm
reflect von Neuman machine — accessing memory and hardware directly. This fined grained control enables the use of low-level optimization techniques \cite{lowLvelOpt}. 
As programmers instruct the machine how to do something, certain performance characteristics and resource constraints can be impelled depending on the context of the computation.
Such techniques encompass memory and cache management, \textit{Instruction-Set} optimization or compiler optimization.

Verbose is missing?

However, while this explicit control provides developers with great functional freedom, it also places a greater burden on them to manage every detail of execution.
Because the programmer must specify each step — from how data is stored in memory to the order in which operations occur — imperative code can become verbose and cluttered obscuring the high-level intent.
Moreover, mutable state and side effects introduce the risk of subtle bugs, especially in concurrent or parallel settings where unintended interactions between state changes
can lead to race conditions. Also, formal reasoning and verification are more complex, since proving correctness requires tracking every state 
transition rather than relying on \textit{referentially transparent} expressions.

Typical use cases for imperative programming include algorithmic or process-oriented problems where an explicit series of steps is natural.
Low-level system programming, embedded development, and performance-sensitive routines frequently rely on imperative constructs to manage hardware resources directly.
In scenarios that demand fine-grained stateful interactions — such as updating user interfaces incrementally, reading and writing files in a specific order, or
controlling physical systems via commands — imperative languages shine by giving developers precise command over execution.
Simulation engines, game logic, and scripting or “recipe”-style automation languages also favor imperatively describing each step to achieve the desired outcome.

\section{Domain-Specific Languages (DSLs)}

Domain-Specific languages are programming or specification languages which are tailored to specific application domain, providing constructs and abstractions
that capture the domain concepts. \cite{whenHowDSL} In stark contrast to general-purpose languages (GLPs) that aim for broad applicability, a DSL waives generality in favor
of higher expressive power in its domain of interest. By offering notation closely related to a domains phrasing, a DSL allows solutions to be formulated at the same level
of abstraction as the domain itself, mapping language primitives onto the domain concepts. Developers, therefore, are able to create programs or specifications more
productively and securely.

The close relationship of DSL and domain furnishes multiple advantages over conventional GLPs: it reduces the amount of general programming knowledge needed, 
expanding the pool of developers from traditional programmers to also domain experts. Moreover, DSLs declutter code, rather than describing a subproblem of the domain
with multiple lines of code, fewer are needed because constructs inherently carry domain specific information, also contributing to broader comprehensibility for those familiar with the domain.
DSLs solutions are also easier to formally verify, as they are restricted to a limited context, than GLPs for which proving correctness is extremely hard.

However, DSLs can suffer from limited scope and flexibility even within their target domain, forcing developers to fall back on general-purpose languages when they need functionality outside the 
DSL's narrow focus. Furthermore, if the domain itself is complex or rapidly evolving, creating and maintaining a DSL may still demand deep domain expertise, because the abstractions must 
accurately capture all necessary domain rules and nuances.

\subsection{Imperative DSLs}

Whichever paradigm a DSL follows is mostly guided by the specifics of the domain to describe. Even though DSLs usually adhere to a declarative approach \cite{sigplanDSL}, which follows from a
major motivation: wanting to describe \textit{what} is expected of the domain in high-level abstraction, a domain may emit qualities that invoke more imperative thinking.
If the nature of a domain is inherently composed of sequential actions, consists of procedures or stateful operations, the imperative paradigm is more appropriate.
Imperative DSLs require the user to script solutions in step-by-step fashion using domain-specific commands and statements. Essentially,the user embeds domain concepts as a sequential programming model,
where the control flow and state transitions of the domain computations have to be managed.

In domains where fine-grained control and explicit ordering matter, imperative DSLs provide maximal flexibility, because every aspect of execution is adjustable.
This level of control becomes crucial when timing or order of operations yield different outcomes, or when fine-tuned performance optimizations are required.
Furthermore, building an imperative DSL is often simpler than creating a declarative one: typically, an interpreters process each command in sequence, 
or a translator converts DSL instructions into a host general-purpose language.

Mybe just leave the below out

Imperative DSLs shine in scripting and macro contexts, where users sequence domain-specific commands to drive behavior, examples include Excel macros that
automate spreadsheet tasks or scripts within a game engine that dictate character actions and game logic. They are also well suited to testing frameworks
and scenario specifications, where user interfaces describe interactions by specifying a timely sequence of user inputs. In robotics and industrial automation, 
imperative DSLs allow practitioners to program custom action sequences—such as moving a robotic arm
to pick up an object, then placing it precisely—where each step must follow a strict order to ensure safety and accuracy. More broadly, any process-control 
domain that relies on precise sequencing and stateful interactions—be it a complex simulation workflow, an interactive operation pipeline, or automated machinery
—benefits from the clarity and control that an imperative DSL provides.
